---
title: | 
  | Introduction to Visualization with R
  | DataViz Meetup
author: "Rob Carnell"
date: "25 Apr 2017"
output:
  beamer_presentation:
    includes:
      in_header: DataVisualizationMeetUpBeamerPreamble.tex
    keep_tex: yes
    df_print: default
---

```{r, echo=FALSE, results=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
# write out the mtcars dataset so that I can read it back in later
write.csv(mtcars, file="mtcars.csv")
# check to ensure all necessary packages are installed
if (!require(knitr)) install.packages("knitr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(maps)) install.packages("maps")
if (!require(hexbin)) install.packages("hexbin")
```

## Goals for the Presentation
### In One Hour...

Provide enough familiarity with R so that you can:

- Install **R** and **RStudio**
- Understand how to get help
- Read a CSV file into **R**
- Change data types
- Create `base` graphics plots
- Create `ggplot2` graphics plots
- Have FUN!

What I won't do:

- This is not a **master class**, only an introduction
- There is sometimes a **better way**.  I'm trying to be clear, not efficient.

## Introduction

### A Roundabout Analytics Career

- B.S. Physics, US Naval Academy
- U.S. Navy, Nuclear Engineer, Submariner
- M.E.M. Engineering Management, Old Dominion University
- M.S. Statistics, Miami University
- Battelle (back when we called this **Statistics**)
- Huntington Bank

### **R** User

- Started with **R** in 2004
- Also develop in C/C++, C#/ASP.NET, Java, python, FORTRAN

## What is **R**?

``\textbf{R} is a language and environment for statistical computing and graphics.'' \url{www.r-project.org}

\begin{itemize}
\item \textbf{R} Installation:  \url{https://cloud.r-project.org/}
\item Install \textbf{RStudio}:  \url{https://www.rstudio.com/}
\end{itemize}

\textbf{R} is Open Source, covered by the GPL, and there are other implementations/copies...

\begin{itemize}
\item Microsoft R Open: \url{https://mran.microsoft.com/open/}
\item \texttt{pqr}: \url{http://www.pqr-project.org/}
\item \texttt{fastr}: \url{https://github.com/allr/purdue-fastr}
\end{itemize}

## **R** is Full of Tricks

\begin{itemize}
\item Extensible with packages
\item \texttt{ROxygen} and \texttt{RUnit} for function decoration and unit testing
\item Self Documenting + \href{https://en.wikipedia.org/wiki/Literate_programming}{Literate Programming}
\begin{itemize}
\item This whole presentation was created in \textbf{R/RStudio} with the \texttt{knitr} package
\end{itemize}
\item Coding in C/C++ using the \texttt{Rcpp} suite of headers
\item Connections to \texttt{Java}, \texttt{python}, \texttt{C\#}, \texttt{HTML}, \TeX
\item Connections to data in SQL Server, Oracle, Hadoop, MySQL, sqlite, JDBC, ODBC, PostgresSQL, various NoSQL...
\end{itemize}

## How to get help
### Help will (almost) always be given on \textbf{R} to those who ask for it.

\href{http://r.789695.n4.nabble.com/R-help-f789696.html}{\textbf{R} Help List}

\begin{itemize}
\item Read \href{https://www.r-project.org/mail.html}{The Posting Guide}
\item Understand \href{https://docs.oracle.com/cd/E19957-01/806-3568/ncg_goldberg.html}{Floating Point Arithmetic}
\item Create a simple, reproducible example that doesn't need all your data
\end{itemize}

\href{https://stats.stackexchange.com/questions/tagged/r}{CrossValidated} (R Stats StackExchange) and 
\href{http://stackoverflow.com/questions/tagged/r}{StackOverflow}

\begin{itemize}
\item A little more forgiving of malformed questions
\end{itemize}

In \textbf{R}:
\begin{itemize}
\item \texttt{?command} e.g. \texttt{?plot}, \texttt{?mean}, \texttt{?lm}
\item \texttt{args(command)} also tab-complete in \texttt{RStudio}
\item \texttt{help.search("command")} also help tab in \texttt{RStudio}
\end{itemize}

## Read a CSV

```{r}
cars_data <- read.csv(file.path("mtcars.csv"), stringsAsFactors=FALSE) 
# also an internal dataset - use ?mtcars
# column names
names(cars_data)
# rename the CSV row names into a column
# negative index means drop that element
names(cars_data) <- c("model", names(cars_data)[-1]) 
head(cars_data)
```

## Change Data Types

```{r}
# data structure
str(cars_data)
# V versus Straight - should be a factor
cars_data$vs <- factor(cars_data$vs, labels=c("V","S"))

# can also use cars_data[["vs"]] or cars_data[,9] 
#   or cars_data[,"vs"] or cars_data[,which(names(cars_data) == "vs")]
# advantage of cars_data$vs is tab-complete
# advantage of cars_data[["vs"]] is that you can code the column name
#   name_of_column <- "vs"
#   cars_data[[name_of_column]]

# with and within allow the data.frame name to not be repeated
cars_data <- within(cars_data, am <- factor(am, labels=c("Automatic","Manual")))
class(cars_data$vs)
```

## Other ways to obtain data

- `readLines`
- `read.table`, `read.delim`, `read.fwf`, `readBin`
- `openxlsx::read.xlsx`
- `RODBC::odbcConnect` and `RODBC::sqlQuery`
- `DBI::dbConnect`, `DBI::dbSendQuery`, `RSQLite`, `RMySQL`

### Note
\begin{itemize}
\item \textbf{R} arrays are 1-based in the interpreter, 0-based in the underlying \texttt{C} code. 
\begin{itemize}
\item It's how statisticians think.  \texttt{C} uses 0-based because of pointer operations.
\end{itemize}
\end{itemize}

# Base Graphics

## Base Graphics

- Internal to R-core, no packages necessary
- Not very object oriented.  Based on S3 object framework.
- Create layers on the graphics device
- e.g. `plot.default`, `plot.lm`, `plot.ecdf`

## Basic Plot Function
### Options

- `x, y = NULL`: x and y coordinates.  y defaults to `NULL` if not supplied
- `type`:  Type of plot.  `p` for point, `l` for line, etc.
- `xlim, ylim`: x and y limits of the plot `c(lower, upper)`
- `log`: log-scale axis
- `main, sub`: main and sub-titles
- `xlab, ylab`: x and y axis labels
- `axes`: should the axes be displayed?
- `...`: see `?par` for other graphical options

```{r}
args(plot.default)
```

## Scatterplot

```{r fig.width=6, fig.height=4}
with(cars_data, plot(x=wt, y=mpg)) # or plot(cars_data$wt, cars_data$mpg)
```

## Scatterplot - more options

```{r fig.width=6, fig.height=4}
# pch = point character, col = color, xlab = x-axis label, main = main title
with(cars_data, plot(x=wt, y=mpg, pch=19, col=gear, xlab="Weight (1000lb)", 
                     ylab="Miles Per Gallon", 
                     main="Relationship of mpg to weight"))
```

## Side Note on ColorBrewer

Prof Brewer at Penn State University \url{http://colorbrewer2.org}

- Great colors for maps and plots!

- Sequential (numeric), diverging (numeric with a transition), and qualitative (factors, categorical)

`require(RColorBrewer)` - Method for adding packages

```{r eval=TRUE, echo=FALSE, message=FALSE}
require(RColorBrewer)
```

## Scatterplot - Add legend, use better colors, change axis

```{r fig.width=6, fig.height=4, message=FALSE}
my_pal <- brewer.pal(5, "Blues") # gear can take values 3, 4, 5
par(mar=c(8,4,1,1))
with(cars_data, plot(x=wt, y=mpg, pch=19, col=my_pal[gear], xlab="Weight (lb)", 
                     ylab="Miles Per Gallon", axes=FALSE))
axis(2); axis(1, at=0:7, labels=(0:7)*1000)
legend("topright", legend=c("3","4","5"), col=my_pal[3:5], pch=19, 
       pt.cex=1.5, title="Number of Gears", box.col=NA)
box()
```

## Simple Regression

$Y_{mpg} = \beta_{0} + \beta_{1} X_{wt} + \beta_{2} X^{2}_{wt} + \epsilon$, $\epsilon \sim N(0,\sigma^2)$
\medskip

Create the **R** model like this:

```{r}
# cars_data$wt_squared <- cars_data$wt^2
# cars_lm <- lm(mpg ~ wt + wt_squared, data=cars_data)
# OR
# cars_lm <- lm(mpg ~ poly(wt, 2, raw=TRUE), data=cars_data)
# OR
cars_lm <- lm(mpg ~ wt + I(wt^2), data=cars_data)
print(cars_lm)
```

## Scatterplot - Add Regression Line

```{r fig.width=6, fig.height=4, message=FALSE}
par(mar=c(8,4,1,1))
with(cars_data, plot(x=wt, y=mpg, pch=19, col=my_pal[gear], xlab="Weight (1000lb)", 
                     ylab="Miles Per Gallon"))
wt_seq <- seq(1, 6, length=100); pred_mpg <- predict(cars_lm, newdata=data.frame(wt=wt_seq))
lines(wt_seq, pred_mpg, lty=2, col="black")
legend("topright", legend=c("3","4","5","model"), col=c(my_pal[3:5],"black"), 
       pch=c(rep(19, 3), NA), lty=c(rep(NA, 3), 2), cex=0.75,
       pt.cex=1.5, title="Number of Gears")
```

## Histogram

```{r fig.height=4, fig.width=6}
hist(cars_data$mpg, xlab="MPG", ylab="Frequency", main="")
```

## Boxplot

```{r fig.width=6, fig.height=4}
with(cars_data, boxplot(mpg ~ am, xlab="", ylab="MPG", boxwex=0.4))
with(cars_data, points(x=am, y=mpg, pch=as.character(vs), cex=0.5))
```

## Explore the Data

```{r fig.width=6, fig.height=4}
pairs(subset(cars_data, select=c("mpg","cyl","disp","hp","wt")))
```

# ggplot2

## ggplot2

- <http://ggplot2.tidyverse.org/index.html>
- Created by Hadley Wickham
- Based on the `Grammar of Graphics`
- Uses the `grid` graphics system
- More object-oriented than `base` graphics
- Requires data in a ``Long'' format. One row per observation.  
- Don't use indicators, use labels instead.
- Many of the plots at the end of the presentation come from the `ggplot2` website

`require(ggplot2)`

```{r eval=TRUE, echo=FALSE, message=FALSE}
require(ggplot2)
```

## ggplot2 - Basic Options

- `data = NULL`: the `data.frame` containing the data
- `mapping = aes()`: the aesthetics of the plot.  Any option defined in the `aes()`
command will turn into an axis or legend
- `...`: any option defined outside `aes()` will not become a legend entry

### aes()

- `x, y, z`
- `alpha`: transparency
- `colour`: line and point color, `fill`: the fill color
- `group`: data groupings
- `shape`: the plotting shape, `size`: the point size
- `linetype`: the type of line (solid, dashed, dotted, etc.)

```{r}
args(ggplot)
```

## ggplot2 - Scatterplot

```{r fig.width=6, fig.height=4}
g <- ggplot(cars_data, aes(x=wt, y=mpg)) + geom_point()
plot(g)
```

## ggplot2 - Scatterplot (cont.)

```{r fig.height=4, fig.width=6}
ggplot(cars_data, aes(x=wt, y=mpg, col=hp)) + geom_point() +
  xlab("Weight (lb)")  + ylab("Miles per Gallon") + 
  scale_x_continuous(label=function(x) 1000*x) + scale_color_distiller() +
  geom_line(aes(x=wt_seq, y=pred_mpg, lty=factor("Model")), 
            data=data.frame(wt_seq, pred_mpg), col="black") +
  labs(col="Power (hp)", lty="")
```

## ggplot2 - Barplot

```{r fig.width=6, fig.height=4}
ggplot(cars_data, aes(as.factor(cyl))) + 
  geom_bar(fill=brewer.pal(3, "Greens")[3], width=0.5) +
  xlab("Cylinders") + ylab("Number of Car Models")

```

## ggplot2 - Barplot (cont)

```{r fig.width=6, fig.height=4}
ggplot(cars_data, aes(cyl, fill=am)) + coord_flip() +
  geom_bar(width=1) + scale_fill_brewer(type="qual", palette="Set2") + 
  xlab("Cylinders") + ylab("Number of Car Models") + labs(fill="Transmission")
```

## ggplot2 - Histogram

```{r fig.width=6, fig.height=4}
ggplot(cars_data, aes(mpg)) + 
  geom_histogram(binwidth=1, fill=brewer.pal(3, "Greens")[3]) +
  xlab("Miles per Gallon") + ylab("Count")
```

## ggplot2 - Boxplot

```{r fig.width=6, fig.height=4}
ggplot(cars_data, aes(x=am, y=mpg)) + geom_boxplot(width=0.5) +
  geom_point(position=position_jitter(width=0.1)) +
  xlab("Transmission") + ylab("Miles per Gallon")
```

## ggplot2 - Dotplot

```{r fig.height=4, fig.width=6}
ggplot(cars_data, aes(x=mpg, group=am)) +
  geom_dotplot(method="histodot", binwidth=2) + xlab("Miles per Gallon") +
  facet_grid(.~am) + scale_y_continuous(NULL, breaks=NULL) 
```

# ggplot2 - Other Great Capabilities

## ggplot2 - Multiple Legends

```{r fig.height=4, fig.width=6}
ggplot(iris, aes(x=Sepal.Length, y=Petal.Length, col=Species, size=Petal.Width)) +
  geom_point(alpha=0.6) + scale_color_brewer(type="qual", palette="Set2") +
  xlab("Sepal Length") + ylab("Petal Length") +
  labs(size="Petal Width", col="Species")
```

## ggplot2 - maps

```{r echo=FALSE, message=FALSE, warnings=FALSE}
require(maps)
crimes <- data.frame(state = tolower(rownames(USArrests)), USArrests)
crimesm <- reshape2::melt(crimes, id = 1)
states_map <- map_data("state")
```

```{r fig.width=6, fig.height=4}
ggplot(crimesm, aes(map_id = state)) +
  geom_map(aes(fill = value), map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat) +
  facet_wrap( ~ variable) + scale_fill_distiller(palette="YlOrRd") + theme_bw() +
  scale_x_continuous(NULL, breaks=NULL) + scale_y_continuous(NULL, breaks=NULL) +
  labs(fill="Number Per\nState")
```

## ggplot2 - ribbon plot

```{r echo=FALSE, message=FALSE, warnings=FALSE}
huron <- data.frame(year = 1875:1972, level = as.vector(LakeHuron), level.se=0)
ts_huron <- ts(unlist(LakeHuron), start=1875, end=1972)
ar_huron <- arima(x=ts_huron, order=c(2,1,3))
huron2 <- data.frame(year = 1973:(1973+10-1), level = predict(ar_huron, n.ahead=10))
names(huron2) <- c("year","level","level.se")
```

```{r fig.height=4, fig.width=6}
ggplot(rbind(huron, huron2), aes(year)) + 
  geom_ribbon(aes(ymin = level - level.se, ymax = level + level.se), fill = "grey") +
  geom_line(aes(y = level), col="steelblue", lwd=1) +
  xlab("Year") + ylab("Lake Huron Water Level") +
  ggtitle("ARIMA prediction of Lake Huron Water Levels (ft)")
```

## ggplot2 - hex density plot

```{r fig.width=6, fig.height=4}
ggplot(diamonds, aes(carat, price)) + geom_hex(binwidth=c(.1,500)) +
  scale_fill_distiller(type="seq", palette="Greens", labels=scales::comma) +
  scale_y_continuous(labels=scales::dollar) +
  xlab("Weight (Carat)") + ylab("Price") + labs(fill="Count") + theme_bw()
```

## ggplot2 - contour plot

```{r fig.width=6, fig.height=4}
ggplot(faithfuld, aes(waiting, eruptions, z=density, col=..level..)) + 
  geom_contour() + xlab("Waiting Time to Next Eruption (min)") +
  ylab("Length of Eruption (min)") + labs(col="Density") +
  ggtitle("Two Clusters of Eruption Types")
```

## Thank-you

**Presentation Code**: <https://github.com/bertcarnell/presentations/tree/master/2017/DataVizMeetUp>

